---
title: "PR1 VD"
author: "Gemma Bargalló Solé"
date: "2025-11-26"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      position: left
    toc_depth: 5   
  word_document: default
  pdf_document: default
---

# 1. MAMMALS DATASET
## 1.1 DATA IMPORT

```{r packages import, warning=FALSE, message=FALSE}
library(DataExplorer)
library(dplyr)
library(ggplot2)
library(gt)
library(stringr) 
```

We start by working with the data on Irish mammals. We observe the first 6 records and generate a summary to see the size and structure of our data.

```{r Mammals data import}
MOI <- read.delim("MammalsOfIreland2016-2025.txt")
head(MOI)
summary(MOI)
str(MOI)
```

The dataset has 14,394 rows (or observations) and 29 columns (or variables).

## 1.2 DATA CLEANSING

We start by performing data cleansing. We review the dataset to detect exact duplicates and format inconsistencies (e.g., casing, data types). We identify null (missing) values and decide whether to impute, remove, or flag them. Finally, we correct or eliminate outliers and invalid values that could distort the analysis.

```{r plot intro}
# We observe a summary plot of the variable types and the missing values (NAs)
plot_intro(MOI, title = "Description of MOI database")
```

We have no complete rows.

### 1.2.1 Duplicates and format inconsistencies

```{r duplicates}
duplicatedMOI <- duplicated(MOI)
duplicatedMOI
sum(duplicated(MOI))
```
There are no duplicate records.

Now, we want to recode the nature of each variable based on its appropriate data type (class).

```{r format inconsistencies}
# First, we convert the date columns
MOI$StartDate <- as.Date(MOI$StartDate, format = "%d/%m/%Y")
MOI$EndDate <- as.Date(MOI$EndDate, format = "%d/%m/%Y")
MOI$Date <- as.Date(MOI$Date, format = "%d/%m/%Y")

# Now we transform the variables (or columns) related to taxonomic groups
MOI$TaxonVersionKey <- as.factor(MOI$TaxonVersionKey)
MOI$TaxonName <- as.factor(MOI$TaxonName)

# Geographical variables
MOI$Projection <- as.factor(MOI$Projection)

# Places
MOI$SiteName <- as.factor(MOI$SiteName)
MOI$County <- as.factor(MOI$County)
MOI$Vice.county <- as.factor(MOI$Vice.county)

# Others
MOI$Recorder <- as.factor(MOI$Recorder)
MOI$Type.of.sighting <- as.factor(MOI$Type.of.sighting)
```


### 1.2.3 Missing values

Let's proceed to examine, count, and treat the missing data (NAs).

```{r whitespaces and missing values, warning=FALSE, message=FALSE}
# Now we replace whitespace with NA

MOI <- MOI %>%
  mutate(
    # 1. Remove extra whitespace, convert multiple spaces to a single one,
    #    and leave empty ("") if there were only spaces.
    #    The trimws() function also works here, but str_squish is very powerful.
    across(
      where(is.character),
      ~ stringr::str_squish(.x)
    )
  ) %>%
  mutate(
    # 2. Convert the resulting empty strings ("") to NA
    across(
      where(is.character),
      ~ na_if(.x, "")
    )
  )


missing_values  <- colSums(is.na(MOI))

# Calculate the percentage of missing values per column
total_values <- nrow(MOI)
percentage_missing <- (missing_values / total_values) * 100

# Convert to dataframe
missing_values_table <- data.frame(
  Variable = names(missing_values),
  Missing_Values = missing_values,
  Percentage_Missing = round(percentage_missing, 2)
)

# Create the table using gt
gt_table <- gt(missing_values_table) %>%
  tab_header(
    title = "Table of missing values by variable"
  ) %>%
  fmt_number(
    columns = vars(Percentage_Missing),
    decimals = 2
  ) %>%
  cols_label(
    Variable = "Variable",
    Missing_Values = "Missing values",
    Percentage_Missing = "% Missings"
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  )

# Show the table
gt_table
```

