---
title: "PR1 VD"
author: "Gemma Bargalló Solé"
date: "2025-11-26"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      position: left
    toc_depth: 5   
  word_document: default
  pdf_document: default
---

# 1. MAMMALS DATASET
## 1.1 DATA IMPORT

```{r packages import, warning=FALSE, message=FALSE}
library(DataExplorer)
library(dplyr)
library(ggplot2)
library(gt)
library(stringr) 
library(forcats)
```

We start by working with the data on Irish mammals. We observe the first 6 records and generate a summary to see the size and structure of our data.

```{r Mammals data import}
MOI <- read.delim("MammalsOfIreland2016-2025.txt")
head(MOI)
summary(MOI)
str(MOI)
```

The dataset has 14,394 rows (or observations) and 29 columns (or variables).

## 1.2 DATA CLEANSING

We start by performing data cleansing. We review the dataset to detect exact duplicates and format inconsistencies (e.g., casing, data types). We identify null (missing) values and decide whether to impute, remove, or flag them. Finally, we correct or eliminate outliers and invalid values that could distort the analysis.

```{r plot intro}
# We observe a summary plot of the variable types and the missing values (NAs)
plot_intro(MOI, title = "Description of MOI database")
```

We have no complete rows and a 13.8% of missings.

### 1.2.1 Duplicates and format inconsistencies

```{r duplicates}
duplicatedMOI <- duplicated(MOI)
duplicatedMOI
sum(duplicated(MOI))
```
There are no duplicate records.

Now, we want to recode the nature of each variable based on its appropriate data type (class).

```{r format inconsistencies, warning=FALSE, message=FALSE}
# First, we convert the date columns
MOI$StartDate <- as.Date(MOI$StartDate, format = "%d/%m/%Y")
MOI$EndDate <- as.Date(MOI$EndDate, format = "%d/%m/%Y")
MOI$Date <- as.Date(MOI$Date, format = "%d/%m/%Y")

# Now we transform the variables (or columns) related to taxonomic groups
MOI$TaxonVersionKey <- as.factor(MOI$TaxonVersionKey)
MOI$TaxonName <- as.factor(MOI$TaxonName)

# Geographical variables
MOI$Projection <- as.factor(MOI$Projection)

# Places
MOI$SiteName <- as.factor(MOI$SiteName)
MOI$County <- as.factor(MOI$County)

```


### 1.2.3 Missing values

Let's proceed to examine, count, and treat the missing data (NAs).

```{r whitespaces and missing values, warning=FALSE, message=FALSE}
# Now we replace whitespace with NA

MOI <- MOI %>%
  mutate(
    # 1. Remove extra whitespace, convert multiple spaces to a single one,
    #    and leave empty ("") if there were only spaces.
    #    The trimws() function also works here, but str_squish is very powerful.
    across(
      where(is.character),
      ~ stringr::str_squish(.x)
    )
  ) %>%
  mutate(
    # 2. Convert the resulting empty strings ("") to NA
    across(
      where(is.character),
      ~ na_if(.x, "")
    )
  )


missing_values  <- colSums(is.na(MOI))

# Calculate the percentage of missing values per column
total_values <- nrow(MOI)
percentage_missing <- (missing_values / total_values) * 100

# Convert to dataframe
missing_values_table <- data.frame(
  Variable = names(missing_values),
  Missing_Values = missing_values,
  Percentage_Missing = round(percentage_missing, 2)
)

# Create the table using gt
gt_table <- gt(missing_values_table) %>%
  tab_header(
    title = "Table of missing values by variable"
  ) %>%
  fmt_number(
    columns = vars(Percentage_Missing),
    decimals = 2
  ) %>%
  cols_label(
    Variable = "Variable",
    Missing_Values = "Missing values",
    Percentage_Missing = "% Missings"
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  )

# Show the table
gt_table
```

We remove those variables that have more than 70% missing values, as they do not provide us with any information.

```{r removing variables and data type correction}
# New dataframe without the columns that have more than 70% missing values.
MOI2 <- MOI %>% 
  select(-SurveyKey, -SampleKey, -SiteKey, -Determiner, -ImagePath, -Habitat.description, -Record.comment, -Life.stage, -Survey.name, -Vice.county)

MOI2$Type.of.sighting <- as.factor(MOI2$Type.of.sighting)

# Final missings
colSums(is.na(MOI2))

```
We will now investigate the remaining missing values in our retained variables to decide on an imputation or treatment strategy.


#### DATE

One of our three date variables contains 29 NAs. Our strategy is to verify the consistency between `Start date` and `End date`—which largely seem identical—to find a pattern for imputing the missings in the primary `Date` variable.

```{r}
identical(MOI2$StartDate, MOI2$EndDate)
```
They aren't identical. 

Let's see which ones aren't identical. 

```{r}
# Identify which rows are different
logical_differences <- (MOI2$StartDate != MOI2$EndDate | 
                        MOI2$StartDate != MOI2$Date | 
                        MOI2$EndDate != MOI2$Date)

# Convert the NAs resulting from the comparison to FALSE (since NAs are neither TRUE nor FALSE for R)
# This prevents subset/filter from removing rows where the difference is NA (e.g., NA != NA)
logical_differences[is.na(logical_differences)] <- FALSE

# Show the row numbers (indices) where the difference is found
print(which(logical_differences))

# 4. Show the dataframe of differences (including indices as row names)
diferences <- MOI2[logical_differences, c("StartDate", "EndDate", "Date")]
print(diferences)
```

The NAs in the `Date` variable occur precisely where `StartDate` $\neq$ `EndDate`, indicating a multi-day sampling event. We can now use this information to impute the missing `Date` values, possibly by using the `StartDate`.

```{r warning=FALSE, message=FALSE}
# Replace the NA in Date with the value from StartDate only when Date is NA.
MOI2 <- MOI2 %>%
  mutate(Date = if_else(is.na(Date), StartDate, Date))

sum(is.na(MOI2$Date))
```

#### GRID REFERENCE

The `GridReference` variable is unnecessary since the complete and non-missing geographic data is already provided by the East and North coordinates, so let's remove it.

```{r}
MOI2 <- MOI2 %>%
  select(-GridReference)
```


#### RECORDER

We substitute the 5 NA records in the `Recorder` variable with the category 'Unknown' and ensure it is of factor type.

```{r}
MOI2$Recorder[is.na(MOI2$Recorder)] <- "Unknown"
MOI2$Recorder <- as.factor(MOI2$Recorder)
```


#### TYPE OF SIGHTING

We substitute the 3936 NA records in the `Type.of.sighting` variable with the category 'Unknown'.

```{r}
MOI2$Type.of.sighting <- MOI2$Type.of.sighting %>%
  fct_na_value_to_level(level = "Unknown")
```


#### ABUNDANCE

We want to see first the categories of `Abundance` variable.

```{r}
MOI2$Abundance <- as.factor(MOI2$Abundance)
summary(MOI2$Abundance)
```


We're going to transform the mixed text and numeric entries in `Abundance` into simplified categories.

```{r}
MOI2$Abundance <- as.character(MOI2$Abundance)
MOI2 <- MOI2 %>%
   mutate(across(where(is.character), ~ str_squish(.x))) %>%
   mutate(across(where(is.character), ~ na_if(.x, "")))


MOI2 <- MOI2 %>%
  mutate(
    Abundance_Clean = str_to_lower(Abundance), # Convert all to lowercase

    Abundance_Simplified = case_when(
      is.na(Abundance) ~ NA_character_,

      # 1. "Zero" or "Absent" Categories
      str_detect(Abundance_Clean, "^0$|cap|none|^na$") ~ "Absent",

      # 2. "Low" Categories (Small numbers, simple text, or less than 10)
      str_detect(Abundance_Clean, "^[1-9]$|^one|^two|^i$|male|scat|pup|young|breed|max\\.|a time") ~ "Low",

      # 3. "High" Categories (Large numbers, high ranges)
      str_detect(Abundance_Clean, "100|207|300|450|500|600|80\\+|\\d{3,}|[5-9][0-9]\\+") ~ "High",

      # 4. "Medium" Categories (Remaining larger numbers, several, 10+)
      TRUE ~ "Medium_or_Other"
    )
  ) %>%
  # Convert the simplified column to an ordered factor
  mutate(
    Abundance_Simplified = factor(Abundance_Simplified,
                                  levels = c("Absent", "Low", "Medium_or_Other", "High"),
                                  ordered = TRUE)
  )
```

We want to recalculate de NA's:

```{r}
# Select only the columns of interest and sum the NAs
MOI2 %>%
  select(Abundance, Abundance_Clean, Abundance_Simplified) %>%
  summarise(
    NA_Abundance_Original = sum(is.na(Abundance)),
    NA_Abundance_Clean = sum(is.na(Abundance_Clean)),
    NA_Abundance_Simplified = sum(is.na(Abundance_Simplified))
  )
```

We are imputing the remaining missing values in the Abundance_Simplified factor by assigning them the level 'Unknown'.

```{r}
MOI2 <- MOI2 %>%
  mutate(
    Abundance_Simplified = Abundance_Simplified %>%
      fct_na_value_to_level(level = "Unknown")
  )

MOI2 <- MOI2 %>%
  select(-Abundance, -Abundance_Clean)
```

Finally, we check that we no longer have any missing values in our dataframe.

```{r}
# Final missings
colSums(is.na(MOI2))
```
Our mammal dataset is ready/prepared.



# 2. PLANTS DATASET
## 2.1 DATA IMPORT

Now we are going to work with the data on Irish Plants. We observe the first 6 records and generate a summary to see the size and structure of our data.

```{r Plants data import}
POI <- read.delim("DiscreteVascularPlantSurveys.txt")
head(POI)
summary(POI)
str(POI)
```

The dataset has 30680 rows (or observations) and 27 columns (or variables).


## 2.2 DATA CLEANSING

We start by performing data cleansing. We review the dataset to detect exact duplicates and format inconsistencies (e.g., casing, data types). We identify null (missing) values and decide whether to impute, remove, or flag them. Finally, we correct or eliminate outliers and invalid values that could distort the analysis.

```{r plot intro}
# We observe a summary plot of the variable types and the missing values (NAs)
plot_intro(POI, title = "Description of POI database")
```

We have no complete rows and a 15% of missings.

### 2.2.1 Duplicates and format inconsistencies

```{r duplicates}
duplicatedPOI <- duplicated(POI)
duplicatedPOI
sum(duplicated(POI))
```
We have 98 duplicated records. 

We are going to filter out the duplicated rows to keep only the unique observations.

```{r}
# Create a new dataframe that contains only unique rows
POI <- POI %>% 
  distinct()

#Check the new size of the dataframe
# The difference between the original number of rows and POI_net should now be exactly 98
nrow(POI)
```
Now, we have 30582 observations.

There are no duplicate records.

We want to recode the nature of each variable based on its appropriate data type (class).

```{r format inconsistencies, warning=FALSE, message=FALSE}
# First, we convert the date columns
POI$StartDate<- as.Date(POI$StartDate, format = "%d/%m/%Y")
POI$EndDate <- as.Date(POI$EndDate, format = "%d/%m/%Y")

# Now we transform the variables (or columns) related to taxonomic groups
POI$TaxonVersionKey <- as.factor(POI$TaxonVersionKey)
POI$TaxonName <- as.factor(POI$TaxonName)

# Geographical variables
POI$Projection <- as.factor(POI$Projection)

# Places
POI$SiteName <- as.factor(POI$SiteName)

```



### 2.2.3 Missing values

Let's proceed to examine, count, and treat the missing data (NAs).

```{r whitespaces and missing values, warning=FALSE, message=FALSE}
# Now we replace whitespace with NA

POI <- POI %>%
  mutate(
    # 1. Remove extra whitespace, convert multiple spaces to a single one,
    #    and leave empty ("") if there were only spaces.
    #    The trimws() function also works here, but str_squish is very powerful.
    across(
      where(is.character),
      ~ stringr::str_squish(.x)
    )
  ) %>%
  mutate(
    # 2. Convert the resulting empty strings ("") to NA
    across(
      where(is.character),
      ~ na_if(.x, "")
    )
  )


missing_values  <- colSums(is.na(POI))

# Calculate the percentage of missing values per column
total_values <- nrow(POI)
percentage_missing <- (missing_values / total_values) * 100

# Convert to dataframe
missing_values_table <- data.frame(
  Variable = names(missing_values),
  Missing_Values = missing_values,
  Percentage_Missing = round(percentage_missing, 2)
)

# Create the table using gt
gt_table <- gt(missing_values_table) %>%
  tab_header(
    title = "Table of missing values by variable"
  ) %>%
  fmt_number(
    columns = vars(Percentage_Missing),
    decimals = 2
  ) %>%
  cols_label(
    Variable = "Variable",
    Missing_Values = "Missing values",
    Percentage_Missing = "% Missings"
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  )

# Show the table
gt_table
```


We remove those variables that have more than 70% missing values, as they do not provide us with any information.

```{r removing variables and data type correction}
# New dataframe without the columns that have more than 70% missing values.
POI2 <- POI %>% 
  select(-SampleKey, -SiteKey, -Determiner, -ImagePath, -Habitat.description, -Abundance)

# Final missings
colSums(is.na(POI2))

```

#### GRID REFERENCE

The `GridReference` variable is unnecessary since the complete and non-missing geographic data is already provided by the East and North coordinates, so let's remove it.

```{r}
POI2 <- POI2 %>%
  select(-GridReference)
```

#### COUNTY 

We substitute the 8883 NA records in the `County` variable with the category 'Unknown'.

```{r}
POI2$County <- POI2$County %>%
  fct_na_value_to_level(level = "Unknown")
POI2$County <- as.factor(POI2$County)

```

