---
title: "pr2"
author: "Gemma Bargalló Solé"
date: "2026-01-07"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      position: left
    toc_depth: 5   
  word_document: default
  pdf_document: default
---

# 1.DATA IMPORT

```{r packages import, warning=FALSE, message=FALSE}
library(readr)
library(here)
library(dplyr)
library(forcats)
library(leaflet)
```

```{r data import}
DB <- read.csv(here("Mammals_Plants_Dataset.csv"))
head(DB)
```

# 2. LITTLE DATA CLEANING

```{r}
# Standardize the Projection column to reflect the current CRS
DB <- DB %>%
  mutate(Projection = "WGS84")

DB$Projection<- as.factor(DB$Projection)
summary(DB$Projection)
```

```{r}
# Change type of some variables 

DB$Date <- as.Date(DB$Date)
DB$StartDate <- as.Date(DB$StartDate)
DB$EndDate <- as.Date(DB$EndDate)
DB$Type <- factor(DB$Type, levels = c("Plant", "Mammal"))
DB <- DB%>%
  mutate(Abundance_Simplified = factor(Abundance_Simplified, 
                                       levels = c("Low", "Medium", "High"), 
                                       ordered = TRUE)) %>%
  mutate(Abundance_Simplified = fct_na_value_to_level(Abundance_Simplified, level = "Unknown"))

```


# 3. DATA VISUALISARION: The spatial distribution and co-occurrence of plant and mammal species across Ireland
## Base structure for an interactive map using R and Leaflet
## 3.1 Minimum data preparation

Before plotting, we create a "map-ready" version of the dataset.

```{r}
DB_map <- DB %>%
  filter(!is.na(lon), !is.na(lat)) %>%
  filter(Type %in% c("Plant", "Mammal"))
```


## 3.2 Defining the color palette

```{r}
pal_type <- c(
  "Plant"  = "#2E8B57",   # green
  "Mammal" = "#2C3E50"    # dark blue
)
```

## 3.3 Base Map initialization

```{r}
map <- leaflet(DB_map) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(lng = -8, lat = 53.3, zoom = 7)

```

## 3.4 Incorporating the plant species layer

```{r}
map <- map %>%
  addCircleMarkers(
  data = DB_map %>% filter(Type == "Plant"),
  lng = ~lon,
  lat = ~lat,
  radius = 4,
  color = pal_type[["Plant"]],
  fillOpacity = 0.65,
  stroke = FALSE,
  popup = ~paste0(
    "<b>Species:</b> ", TaxonName, "<br>",
    "<b>Group:</b> ", Type, "<br>",
    "<b>County:</b> ", County, "<br>",
    "<b>Date:</b> ", Date
  ),
  group = "Plants"
)


```

## 3.5 Incorporating the mammal species layer

```{r}
map <- map %>%
  addCircleMarkers(
    data = DB_map %>% filter(Type == "Mammal"),
    lng = ~lon,
    lat = ~lat,
    radius = 4,
    color = pal_type[["Mammal"]],
    fillOpacity = 0.65,
    stroke = FALSE,
    popup = ~paste0(
      "<b>Species:</b> ", TaxonName, "<br>",
      "<b>Group:</b> ", Type, "<br>",
      "<b>County:</b> ", County, "<br>",
      "<b>Date:</b> ", Date
    ),
    group = "Mammals"
  )

```


## 3.6 Layer control

```{r}
map <- map %>%
  addLayersControl(
    overlayGroups = c("Plants", "Mammals"),
    options = layersControlOptions(collapsed = FALSE)
  )

```

```{r}
map

```

