---
title: "pr2"
author: "Gemma Bargalló Solé"
date: "2026-01-07"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      position: left
    toc_depth: 5   
  word_document: default
  pdf_document: default
---

# 1.DATA IMPORT

```{r packages import, warning=FALSE, message=FALSE}
library(readr)
library(here)
library(dplyr)
library(forcats)
library(leaflet)
library(crosstalk)
```

```{r data import}
DB <- read.csv(here("Mammals_Plants_Dataset.csv"))
head(DB)
```

# 2. LITTLE DATA CLEANING

```{r}
# Standardize the Projection column to reflect the current CRS
DB <- DB %>%
  mutate(Projection = "WGS84")

DB$Projection<- as.factor(DB$Projection)
summary(DB$Projection)
```

```{r}
# Change type of some variables 

DB$Date <- as.Date(DB$Date)
DB$StartDate <- as.Date(DB$StartDate)
DB$EndDate <- as.Date(DB$EndDate)
DB$Type <- factor(DB$Type, levels = c("Plant", "Mammal"))
DB <- DB%>%
  mutate(Abundance_Simplified = factor(Abundance_Simplified, 
                                       levels = c("Low", "Medium", "High"), 
                                       ordered = TRUE)) %>%
  mutate(Abundance_Simplified = fct_na_value_to_level(Abundance_Simplified, level = "Unknown"))

```


# 3. DATA VISUALISARION: The spatial distribution and co-occurrence of plant and mammal species across Ireland
## Base structure for an interactive map using R and Leaflet
## 3.1 Minimum data preparation

We will use a single SharedData object. This ensures that all filters (time and species type) act on the same data stream, preventing synchronization errors.

```{r}
# 1. Ensure Year is a clean integer and Date is correctly formatted
DB_map <- DB %>%
  filter(!is.na(lon), !is.na(lat), !is.na(Date)) %>%
  filter(Type %in% c("Plant", "Mammal")) %>%
  mutate(Year = as.integer(format(as.Date(Date), "%Y")))

# 2. Create a single SharedData object for all interactive elements
shared_data <- SharedData$new(DB_map)
````


## 3.2 Defining the color palette

```{r}
# Define colors for the biological groups
pal_species <- colorFactor(
  palette = c("#2E8B57", "#2C3E50"), 
  domain = DB_map$Type
)
```


## 3.3 Dashboard assembly: Slider and Checkboxes

Using bscols, we will place the time slider and the species selector (checkboxes) at the top of the map. This provides a professional "dashboard" feel.

```{r}
# Create the interactive dashboard
bscols(
  widths = c(12),
  # 1. Time Slider (linked to shared_data)
  filter_slider("year_selector", "Select Observation Year", shared_data, ~Year, width = "100%"),
  
  # 2. Species Selector (replaces the Leaflet layer control for better compatibility)
  filter_checkbox("type_selector", "Select Species Group", shared_data, ~Type, inline = TRUE),
  
  # 3. The Map
  leaflet(shared_data) %>%
    addProviderTiles(providers$CartoDB.Positron) %>%
    setView(lng = -8, lat = 53.3, zoom = 7) %>%
    addCircleMarkers(
      lng = ~lon, lat = ~lat,
      radius = 4,
      color = ~pal_species(Type),
      fillOpacity = 0.65,
      stroke = FALSE,
      popup = ~paste0(
        "<b>Species:</b> ", TaxonName, "<br>",
        "<b>Group:</b> ", Type, "<br>",
        "<b>Year:</b> ", Year
      )
    ) %>%
    addLegend(
      position = "bottomright",
      pal = pal_species,
      values = ~Type,
      title = "Species Group"
    )
)

```




# 4. VISUALIZATION ANSWERS

The main visualization focuses on exploring the spatial distribution of plant and mammal species across Ireland. Through interactive mapping, users can identify regions with higher species richness, observe areas where plants and mammals co-occur, and explore individual species records in detail. Additional contextual variables such as sampling method and observation date allow for a more nuanced interpretation of spatial patterns and potential biases in species detection.


